---
title: "Voice Analytics Workshop"
#author: Christian Hildebrand
format: 
  html:
    theme: flatly
    code-background: true
    #code-fold: true
    code-copy: true
    toc: true
    toc-depth: 3
    toc-location: left
    #toc-title: VAW
---

#### Author & Contact:
Christian Hildebrand <br>
Full Professor of Marketing Analytics <br>
Institute of Behavioral Science & Technology <br> 
University of St. Gallen <br><br>
Email: [christian.hildebrand@unisg.ch](mailto:christian.hildebrand@unisg.ch)
<br>Web: [IBT-HSG Profile](https://www.ibt.unisg.ch/team/christian-hildebrand/) 
<br>Social: [LinkedIn Profile](https://www.linkedin.com/in/christian-hildebrand/)
<br> 
<br> 

```{r setup}
#| echo: true
#| label: setup
#| include: false
#Define path where you have the recordings inside root.dir argument
knitr::opts_knit$set(root.dir = "voiceData/1_alexa") 
```


## Using Voice Analytics to Understand User Frustration 

In this tutorial, we will analyze a short video featuring a female Scottish user attempting to issue a command to Amazon Alexa to play a song on Spotify. This viral video, though amusing, highlights a common frustration many users encounter when trying to communicate effectively with voice-controlled interfaces.

{{< video voiceData/1_alexa/alexa_video.mp4 >}}

For our voice feature analysis, we begin by extracting the audio file from the previous video and converting it into the Waveform audio format. Our focus is centered on two main parts of this interaction:

1.  **Speech Formation of the Wakeword "Alexa"**

2.  **Vocal Changes During the Issuance of a Command ("Alexa, play something is cooking in my kitchen on Spotify by Dana")**

To facilitate our analysis, we first edit the voice recordings, retaining only the segments containing the three consecutive wakewords and the two subsequent user commands. It is noteworthy that the third utterance of "Alexa" was not associated with a command but rather followed by a barrage of profanity and abusive language. In what follows, we further process, edit, and analyze those parts of the interaction. You can download the files for this example [here](https://www.dropbox.com/scl/fo/g689dlmuidyhvgycawkw2/h?rlkey=kfhtnwbq4zahu6mt5w5fmklr6&dl=0).

Our analytical will first leverage the `seewave` package, which has emerged as the gold standard for bioacoustics and audio analysis in R. This versatile package encompasses an impressive array of 130 functions designed for the analysis, manipulation, representation, editing, and synthesis of time-based audio waveforms. While `seewave` serves as a baseline in this introduction, we also make reference to other main packages, such as `tuneR`, `soundgen` and `phonTools`, for their specialized functionalities as needed and our own `voiceR` package that you can find [here](https://cran.r-project.org/web/packages/voiceR/).

## Data Acquisition and Processing

### Reading Sound Files

As previously mentioned, the primary focus of this tutorial centers around the utilization of the seewave package. While it is important to note that `seewave` lacks native capabilities for sound file reading, we overcome this limitation by using functions from complementary packages. It's important to note that different packages use different classes of sound objects, each optimized for specific sound manipulation tasks. Consequently, when choosing an alternative package to load sound data make sure you understand which sound object is created by this package by calling the sound object after you've created it (more below).

In the context of `seewave`, its core functionality is primarily tailored to work  with sound objects of the `Wave` class. These `Wave` class sound objects are created by `seewave` calling the `tuneR` package. Hence, when you work with `seewave`, I recommend generally to use `tuneR` for loading sound data to avoid conflicts with other packages.

The process of loading the two user commands from the interaction with Amazon Alexa and saving these sound files as new objects, `cmd1` and `cmd2`, is accomplished as follows:

To begin, we use the `readWave()` function from the `tuneR` package. This function reads a sound file from a specified location. Here is the step-by-step procedure for loading and saving the user's interaction commands:

```{r Loading sound with tuneR}
#| warning: false

library(tuneR)
cmd1 <- readWave("alexa_cmd1.wav")
cmd2 <- readWave("alexa_cmd2.wav")
```

After we created those objects, let's take a look at the underlying structure of these objects. This output not only underscores the seamless integration of the `wave` class objects generated by the `tuneR` package but also provides key insights into their fundamental characteristics. These characteristics cover:

1.  **Number of Samples**: This indicates the total count of discrete data points in the audio waveform.

2.  **Duration (in seconds)**: The elapsed time in seconds, capturing the length of the audio file.

3.  **Sampling Rate (in Hertz)**: Denoting the rate at which individual samples are taken per second.

4.  **Number of Channels**: Indicates whether the audio is mono (single channel) or stereo (two channels).

5.  **Bit Rate**: Representing the number of bits processed per unit of time.

Calling the two sounds objects shows that they share identical sampling rates, channel numbers, and bit rates. However, we already see differences in duration, with the second sound file being 820 milliseconds longer than the first, something we further explore later when testing specific differences at the feature level.

```{r Calling cmd1 and cmd2}
cmd1
cmd2
```

The `readWave()` function offers additional parameters such as the `from` and `to` arguments, which allow you to select only specific segments within the audio file. By default, these parameters operate in samples units, defining the segment based on sample counts. However, the `readWave()` function also allows you to specify the `units` argument, giving you more flexibility to modify the units of `from` and `to` to seconds, minutes, or even hours.

For example, suppose we want to load only the initial 0.5 seconds and the segment from 0.5 seconds to 2 seconds from the audio file and store them in separate objects, denoted as `cmd1.s1` and `cmd1.s2`, respectively. To do this, we simply have to adjust the `from`, `to`, and `units` arguments as follows:

```{r Specifying preferred section}

cmd1.s1 <- readWave("alexa_cmd1.wav",from=0,to=0.5,units="seconds")
cmd1.s2 <- readWave("alexa_cmd1.wav",from=0.5,to=2,units="seconds")

```

### Playing Sound Files

Sound analysis is an iterative process, requiring frequent listening to parts of a soundwave. Although, R itself cannot play sound files the `seewave`'s `listen( )` function allows us to call the default audio player of the user's operating system. Thus, we could play the previously loaded cmd1 variable by using the `listen()` function.

::: {.callout-note}
Mac users might have to run the following code snippet to access the correct audio driver for tuneR. Simply run the following code if you get a "Permission denied" error notification. Use this code to fix it:
`tuneR::setWavPlayer('/usr/bin/afplay')`.
:::


```{r Playing sound using listen from seewave }
library(seewave)
listen(cmd1)
```

Much like the `readWave()` function, the `listen()` function offers analogous `from` and `to` arguments to precisely determine the sections of interest for auditory playback. Furthermore, it introduces an `f` argument, providing the capability to manipulate the sampling frequency rate. This adjustment allows for the alteration of the original sound's pitch, facilitating the creation of both higher and lower-pitched auditory renditions. The code snippet below demonstrates this transformative process:

```{r play cmd1 sound with 1.1 times higher sampling frequency}
#| eval: false
listen(cmd1, f=cmd1@samp.rate*1.1)
```

```{r play cmd1 sound with 1.1 times lower sampling frequency}
#| eval: false
listen(cmd1, f=cmd1@samp.rate/1.1)
```

### Editing Sound Files

In certain applications, it becomes necessary to perform additional edits on voice files. These edits could include tasks such as (1) extracting specific segments of a soundwave for in-depth analysis, (2) eliminating a series of utterances from a soundwave, (3) trimming periods of silence at the beginning or end of a sound file, and (4) filtering out all unvoiced frames from a sound file.

Both the `tuneR` and `seewave` packages offer a suite of functions designed to address these various editing procedures:

1.  `extractWave( )`: This function facilitates the extraction of desired segments from a soundwave. Users can specify the segments using the `from` and `to` arguments, which we discussed earlier. The `extractWave()` function defaults to samples as the unit, but this can be adjusted using the 'xunit' argument.

2.  `deletew( )`: To remove specific portions from a soundwave, the `deletew()` function is employed. By default, it operates in units of time.

3.  `noSilence( )`: This function is particularly useful for removing periods of silence from the beginning or end of a sound file. Users can fine-tune this process by specifying the desired silence level.

4.  `zapsilw( )`: To eliminate all unvoiced frames from a sound file, the `zapsilw()` function comes into play. Users can tailor this operation by setting the 'threshold' argument, which measures the amplitude threshold in percent distinguishing silence from signal. Additionally, the `zapsilw()` function offers the default feature of generating oscillograms for both the original sound file and the modified version post unvoiced frame removal, providing visual insight into the process.

These functions empower users to efficiently manipulate sound files, ensuring they are tailored to meet the specific requirements of their analyses. To illustrate their practical utility, let's delve into some illustrative examples.

As an instance, consider the application of the `extractWave()` function to pinpoint the initial 700 milliseconds of the soundwave residing in `cmd1`.

```{r Example of (1)}
#Extract first 700ms
cmd1.xtr <- extractWave(cmd1, from = 0, to = 0.7, xunit = "time") 

```

Here, the `from` and `to` arguments are employed to specify the time range of interest. In this case, we isolate the first 700 milliseconds, facilitating a focused analysis.

Alternatively, instead of extracting this segment, we can achieve its removal using the `deletew()` function:

```{r Example of (2)}
#Delete first 700ms
cmd1.rem <- deletew(cmd1, from=0, to=0.7, output="Wave") 
```

Continuing with our exploration of sound manipulation, we can also replicate the first 700 milliseconds threefold with the `repw()` function:

```{r Repeat sound}
#Repeat first 700ms three times
cmd1.rep3 <- repw(cmd1, f=cmd1@samp.rate, times=3, output="Wave")

```

To refine the audio data further, we may opt to remove solely the unvoiced segments at the beginning and end using the `noSilence()` function:

```{r Example of 3}
#Remove only unvoiced start and ending
cmd1.cut <- noSilence(cmd1)
cmd1.cut
cmd1
```

In this particular instance, a comparison between the original sound `cmd1` and the processed version `cmd1.cut`, where we have removed unvoiced segments from the beginning and end, reveals no discernible alteration. This lack of change stems from the fact that `cmd1` initially did not contain any unvoiced segments at either its outset or conclusion.

Alternatively, we could use the `zapsilw()` function to cleanse all the unvoiced elements of `cmd1`:

```{r Example of (4)"}
#| label: fig-oscillogram
#| fig-cap: "Oscillograms of First and Original Command (Upper Panel) and with voice breaks removed (Lower Panel)"
#Remove all unvoiced frames of a soundwave
cmd1.nosil <- zapsilw(cmd1, threshold=1, output="Wave")
cmd1
cmd1.nosil

```

In this case, a close examination of @fig-oscillogram reveals that the `zapsilw()` function has been remarkably successful in eliminating all the unvoiced areas within the soundwave. This process has resulted in a cleaner and more refined audio representation.

Furthermore, when we compare `cmd1` to `cmd1.nosil`, it becomes evident that the duration of the latter is noticeably shorter, clocking in at almost 1 second less. This reduction in duration underscores the effective removal of unvoiced segments, affirming the utility of this process in streamlining the audio data while preserving its crucial vocal elements.

### Writing Sound Files

Following sound file editing, it is often essential to preserve the revised version for future use. To accomplish this, the seewave package offers the convenient `savewav( )` function, designed explicitly for storing R sound objects as .wav files. The process involves specifying the following parameters:

1.  **R Sound Object**: As the first argument, designate the R sound object that you intend to save as a .wav file.

2.  **Sampling Frequency** (`f`): Next, specify the sampling frequency to be associated with the saved .wav file.

3.  **Filename** (`filename`): Finally, provide the desired filename under which the edited sound object will be stored.

It's worth noting that if the sampling frequency is not explicitly defined, the seewave package will automatically utilize the same sampling frequency as the edited R object, simplifying the saving process.

As an illustrative step, let's go ahead and save the modified version of `cmd1`, where all unvoiced segments have been successfully removed, as a .wav file within our system:

```{r save sound file}
savewav(cmd1.nosil, filename = "cmd1_noSilence.wav")  	
```

## Visualizing Sound

Now that we can work with audio files, let's visualize them to better understand what's happening in the voice recordings. Sound visualizations convert audio signals into graphical representations that we can analyze.
We'll focus on three key aspects of sound:

- Amplitude: the loudness or intensity of the sound
- Frequency: the pitch components in the sound
- Time: how these elements change throughout the recording

We'll use two main visualization types:

- Oscillograms: show amplitude changes over time (like waveforms)
- Spectrograms: display frequency and amplitude together over time

These visualizations will help us identify patterns in the user's voice that indicate emotional changes, particularly signs of frustration with Alexa.

### Visualizing Amplitude

Oscillograms offer a visual representation of the instantaneous amplitude of a soundwave plotted against time. They are often referred to as waveforms, as they graphically depict the variations within the sound wave itself. Oscillograms serve as valuable tools for detecting potential changes in loudness over time within a soundwave. In R, you can create oscillograms using the `oscillo()` function from the seewave package. This function requires just one argument, the sound object. Moreover, `oscillo()` provides the flexibility to customize various visual aspects, such as the title (using the `title` argument), label color (via the `collab` argument), and wave color (by setting the `colwave` argument). Additionally, you can specify the `from` and `to` arguments, similar to what we did during data processing, to generate an oscillogram for a specific time interval in seconds.

To gain insights from the oscillograms of the two Alexa commands, we can first visualize the entire soundwave and then zoom in to focus specifically on the articulation of the wakeword.

To plot all four graphs within a single plotting region, we partition the plot into four distinct sections using the standard `par` and `mfrow` arguments in R. To display the wakeword, we make use of the `from` and `to` arguments within the `oscillo()` function.

```{r Oscillogram for each command}
#| label: fig-oscillograms
#| fig-cap: "Oscillograms of First and Second Command (Upper Panel) and Wakeword (Lower Panel)"
#| echo: false
par(mfrow=c(2,2))

oscillo(cmd1, colwave="black", collab="black", title = TRUE)

oscillo(cmd2, colwave="black", collab="black", title = TRUE)

oscillo(cmd1, from=0, to=.7, colwave="blue", collab="red", 
             title = "First     Command - Wakeword Only")

oscillo(cmd2, from=0, to=.7, colwave="blue", collab="red", 
             title = "Second Command - Wakeword Only") 

```

Upon closer examination, as depicted in @fig-oscillograms, a few notable observations come to light:

1.  **Difference in Duration**: It becomes clear that the second command is slightly longer than the first. Remarkably, both commands share identical content, both stating "Alexa, play by Dana." The oscillograms reveal the underlying reason: in the second command, there are longer vocal breaks, representing the time gaps between each spoken word.

2.  **Emphasis on Individual Words**: Notably, there are striking variations in the emphasis placed on individual words when issuing the two commands. For instance, in the lower panel of @fig-oscillograms, we can see the pronounced differences in overall amplitudes. Particularly, the word "Alexa" in the second command exhibits significantly greater amplitude compared to its counterpart in the first command. This observation suggests that, without needing to listen to the voice files or comprehend the spoken content, we can already infer that the second command tends to be louder overall than the first. Such a distinction may hint at potential heightened emotions such as anger, stress, or irritability of the speaker.

### Visualizing Fundamental Frequency

The graphical representation of the fundamental frequency over time, often referred to as an f0 contour or pitch track, holds significant value in acoustic analysis. This visualization unveils essential aspects, including the speaker's fundamental frequency range, pitch variations throughout speech, distinctions between voiceless and voiced segments, as well as patterns of regular and irregular phonation.

In the context of the `seewave` package, the `fund()` function allows us to estimate the fundamental frequency of an R object, employing a short-term cepstral transform, and automatically generates a visual plot of the fundamental frequency.

To demonstrate the practical application and insights derived from this analysis, we narrow our focus to the wakewords uttered by the speaker during their interaction with Amazon Alexa. It's worth noting that although the user issued only two complete commands, the interaction involved three distinct wakewords. The third wakeword, did not lead to the issuance of a command but instead featured the use of offensive language directed at Alexa.

To proceed with this exploration, we initiate the process by reading these distinct wakewords using the `readWave()` function:

```{r load the three wakewords}
w1 <- readWave("alexa_wakeword_1.wav")
w2 <- readWave("alexa_wakeword_2.wav")
w3 <- readWave("alexa_wakeword_3.wav")
```

We can already see clear differences in the durations of these wakewords. Specifically, they span 0.7 seconds, 0.74 seconds, and 1.05 seconds, respectively, highlighting the temporal shift.

With the three wakewords now successfully imported into R, we proceed to concatenate them into a single soundwave using seewave's `bind()` function. We undertake this concatenation to facilitate a comprehensive visual evaluation of how the fundamental frequency evolves from one command to another.

```{r concatenate wake words}
wake_all <- bind(w1,w2,w3)
listen(wake_all)
```

After combining all three wakewords into a single audio file (`wake_all`), we can analyze how the fundamental frequency (pitch) changes across the utterances in the below @fig-pitch. The plot shows the fundamental frequency (pitch) measurements over time, with each circle representing a detected pitch point. The x-axis shows time in seconds (0-2.5s), covering all three wakewords in sequence. The y-axis shows frequency in kHz (0-20 kHz).

Two observations stand out:

- *Rising trend in pitch*: The red dashed line shows a linear model fitted to the data, revealing a clear upward trend in the speaker's fundamental frequency across the three wakewords. This increasing pitch is a strong indicator of growing emotional intensity and frustration.

- *Frequency clusters*: Notice the clustering of points around 0.5s (first wakeword), 1.0-1.5s (second wakeword), and 1.8-2.2s (third wakeword). The points generally move higher in frequency as we progress through the utterances.

This analysis quantitatively confirms what we might hear qualitatively - the user's voice becomes progressively higher-pitched as they repeatedly try to activate Alexa without success. 

```{r plot of fundamental frequencies and trend}
#| label: fig-pitch
#| fig-cap: "Increasing Fundamental Over Time from First, Second, to Third Wakeword"
ff <- fund(wake_all,threshold=1)
mod <- lm(ff[,2]~ff[,1])
abline(mod,col="red",lwd=2,lty=2)
```

To enhance our analysis further, we employ the `formanttrack()` function from the `phonTools` package. This function serves as a practical tool to track the distribution of acoustic energy across various frequency bands, typically in 1000Hz intervals. It's worth noting that the phonTools package exclusively supports mono audio. Therefore, when working with stereo files, as in this scenario, only one of its channels should be specified as an argument. The channel of a wave object can be accessed using the `@` operator along with the name of the desired channel (`left` or `right`). Additionally, since `phonTools` is not specifically designed to handle wave objects, the sampling frequency of the sound file must be manually set using the `fs` argument.

@fig-formanttrack illustrates the first three frequency bands for each wakeword, respectively. These displays complement our earlier observations, illustrating the frequency escalation from the first to the third wakeword, along with an overall increase in variability over time. Given the strong correlation between heightened frequency levels and experiences of stress, anger, or frustration, these visualizations further underscore the extended duration during which the user encountered difficulties in their interaction with Amazon Alexa.

```{r Formants at different time points}
#| warning: false
#| label: fig-formanttrack
#| fig-cap: "Visualizing the First (Black), Second (Red), and Third (Green) Formants Across Wakewords"

library(phonTools)
par(mfrow=c(1,3))
formanttrack(w1@left, fs=w1@samp.rate, formants=3, periodicity=.5)
formanttrack(w2@left, fs=w2@samp.rate, formants=3, periodicity=.5)
formanttrack(w3@left, fs=w3@samp.rate, formants=3, periodicity=.5)

```

### Spectrograms

Spectrograms provide a rich, multi-dimensional portrayal of a soundwave, offering insights into its composition. In this representation, time unfolds along the x-axis, frequency extends along the y-axis, and a third dimension portrays amplitude levels (loudness) through varying color codes. The `seewave` package equips us with the `spectro()` function, which constructs a spectrographic visualization of a time wave. This versatile function demands only a time wave object, such as a sound object, as its input.Furthermore, the `spectro()` function offers several customization options to tailor the appearance of the spectrogram. For instance, the `flim` argument permits us to specify the minimum and maximum frequencies displayed. Additionally, `osc` introduces an oscillogram at the bottom of the spectrogram plot, while `dBref` allows us to define a reference value for the dB range of amplitude.

@fig-spectrograms provides a more nuanced understanding, summarizing the significant distinctions between the initial two commands and the subsequent three consecutive wakewords spoken by the user. Across all spectrograms, three noteworthy points emerge:

1.  **Increase in Loudness**: There is a noticeable escalation in power or loudness, depicted by the deepening reddish color gradients.

2.  **Prolonged Voice Breaks**: A trend of longer voice breaks becomes evident from the first command to the second.

3.  **Frequency Stability**: The first command exhibits relatively stable frequency, whereas the second command displays a moderate upward trend.

Collectively, these observations suggest that the heightened frequency, amplified loudness, and extended voice breaks likely mirror the speaker's experiences of tension, anger, stress, and frustration during their interaction with Amazon Alexa.

```{r spectrogram}

spectro(cmd1, osc=TRUE, flim=c(0,6), dB = "max0", dBref = 2*10e-5)

spectro(cmd2, osc=TRUE, flim=c(0,6), dB = "max0", dBref = 2*10e-5)

spectro(wake_all, osc=TRUE, flim=c(0,6), dB = "max0", dBref = 2*10e-5) 

```


## Acoustic Feature Extraction

Now that we've learned how to visualize sound data, let's extract specific acoustic features that can quantify the patterns we observed. These features are the building blocks of voice analytics and can be used as variables in statistical models or as indicators of emotional states.
We'll organize our analysis into four key domains, each capturing different aspects of the voice signal:

- Time domain: measures related to duration and timing
- Amplitude domain: features related to loudness and energy
- Frequency domain: pitch characteristics and variations
- Spectral domain: voice quality and acoustic perturbations

For each domain, we'll use functions from the `seewave` and `soundgen` packages to extract meaningful metrics that help us understand the user's emotional state.

### Time Domain

The most fundamental measure in the time domain is the duration, typically expressed in seconds or milliseconds, which quantifies the temporal dimension of a soundwave. The `duration()` within the `seewave` package offers a straightforward function to extract overall time measured in seconds. When we apply this function to the two commands, we see that the first command has a duration of `r round(duration(cmd1), 2)` seconds, while the second command spans `r round(duration(cmd2), 2)` seconds.

```{r Computing duration for two commands}
duration(cmd1)
duration(cmd2)
```

Similarly, when we apply the same procedure to the wakewords, we find that they have durations of `r round(duration(w1), 2)` seconds, `r round(duration(w2), 2)` seconds and `r round(duration(w3), 2)` seconds, respectively.

```{r Computing duration for three wakewords}
duration(w1)
duration(w2)
duration(w3)
```

The `soundgen` package includes the **`analyze()`** function, which provides the ability to extract several features across the time, amplitude, frequency, and spectral domain respectively. For example, using this function we can directly extract the number of voiced and unvoiced frames of a sound object. To do so, the results from the `soundgen` function should be stored in an R object. This object results in a list of two data.frames: a detailed data.frame (you can retrieve it by using **`$detailed`**) in which each row represents a Short-Time Fourier Transform frame and each column represents a vocal feature and a summarized `data.frame` (you can retrieve it by using **`$summary`**).

Thus, we can proceed to use the `analyze()` function to extract several vocal features from the first and second commands:

```{r load soundgen and analyze}
#| message: false
#| warning: false
library(soundgen)
feat_cmd1 <- analyze("alexa_cmd1.wav", plot = F)
feat_cmd2 <- analyze("alexa_cmd2.wav", plot = F)
```

Since the summary object represents a data.frame, we can access its columns using the `$` operator followed by the column name. Hence, in order to check which proportion of frames are voiced, we can directly call the `voiced` column from the output extracted by the `analyze()` function:

```{r}
#Returns the proportion of voiced samples
feat_cmd1$summary$voiced
feat_cmd2$summary$voiced
```

Revealing a a greater percentage of vocal breaks in the second (`r round(feat_cmd1$summary$voiced * 100, 2)`%) compared to the first (`r round(feat_cmd2$summary$voiced * 100, 2)`%) command.

### Amplitude Domain

The amplitude of a soundwave dictates its power or loudness, with smaller amplitudes indicating softer sounds and larger amplitudes indicating louder ones. It measures how far air particles deviate from their equilibrium position. To calculate the amplitude at various points in time, we can utilize the `oscillo` function from the `seewave` package. This provides us with various statistics, such as the maximum and minimum amplitudes. In the code snippet below, we can see that the first command generally had a higher volume compared to the second.

```{r max and min amplitude using oscillo}
min(oscillo(cmd1, plot = F))
max(oscillo(cmd1, plot = F))
min(oscillo(cmd2, plot = F))
max(oscillo(cmd2, plot = F))
```

Another method to estimate soundwave amplitude involves computing the root-mean-squared (RMS) of the amplitude envelope. This can be accomplished by combining the `rms()` function to calculate the RMS and the `env()` function to calculate the envelope. This approach provides us with insight into the sound object's average loudness, revealing again that, on average (after removing silent portions), the first command was louder than the second.

The Root Mean Square (RMS) is calculated as follows:

$$RMS = \sqrt{\frac{1}{N}\sum_{n=1}^{N} x_i^2}$$ Here, $x_i$ represents each amplitude envelope point and $N$ represents the total number of points.

```{r amplitude statistics}
#| label: fig-envelope
#| fig-cap: "Amplitude envelope of Command 1 (right) and Command 2 (left)"

par(mfrow=c(1,2))
rms(env(zapsilw(cmd1, plot = F),f=cmd1@samp.rate))
rms(env(zapsilw(cmd2, plot = F),f=cmd1@samp.rate))
```

Moreover, the `analyze()` function from the `soundgen` package also captures a subjective unit of loudness measured in 'sone'. This measure is stored in the 'loudness' column within the data.frame produced by `analyze()`. Using the `mean()` function in R on the `loudness` column computes the average loudness in sone, reaffirming our earlier observations that the second command was generally louder than the first.

```{r Perceived loudness in sone}
feat_cmd1$summary$loudness_mean
feat_cmd2$summary$loudness_mean
```

Additionally, `soundgen` provides the `getLoudness()` function, offering a visual representation of loudness spectrum for soundwaves. It presents a grayscale visualization of the spectrum, with darker areas indicating higher sound pressure levels, and thus, higher loudness.

```{r getLoudness method}
loudness1 <- getLoudness("alexa_cmd1.wav", plot = TRUE)
loudness2 <- getLoudness("alexa_cmd2.wav", plot = TRUE)
```


### Frequency Domain

In this section we focus on the extraction of two main features in the frequency domain: (1) pitch of soundwave and (2) the extent of variability in frequency of the sound-wave. Both of them can be directly extracted from the `summary` of the results extracted from `soundgen`'s `analyze( )` function: (1) `pitch_mean` and (2) `pitch_sd`. Consistent with our expectation, we find a strong increase from `r round(feat_cmd1$summary$pitch_mean, 2)`Hz to `r round(feat_cmd2$summary$pitch_mean, 2)`Hz in the average pitch. Thus, the extracted features confirm the "shrill" and aroused sound of the user's voice after the repeated failure from the first to the second voice command.

```{r mean and deviation of fundamental frequency}
feat_cmd1$summary$pitch_mean
feat_cmd2$summary$pitch_mean

feat_cmd1$summary$pitch_sd
feat_cmd2$summary$pitch_sd

```

To get into a higher level of granularity, we can apply those same analysis only to the wakeword "Alexa". We can see how these key differences are even more prominent: `r round(analyze("alexa_wakeword_1.wav", plot = F)$summary$pitch_mean,2)`Hz on the first wakeword, `r round(analyze("alexa_wakeword_2.wav", plot = F)$summary$pitch_mean, 2)`Hz on the second and `r round(analyze("alexa_wakeword_3.wav", plot = F)$summary$pitch_mean, 2)`Hz on the third.

```{r mean and deviation of wakeword frequency}
feat_w1 <- analyze("alexa_wakeword_1.wav", plot = F)
feat_w2 <- analyze("alexa_wakeword_2.wav", plot = F)
feat_w3 <- analyze("alexa_wakeword_3.wav", plot = F)

feat_w1$summary$pitch_mean
feat_w2$summary$pitch_mean
feat_w3$summary$pitch_mean

feat_w1$summary$pitch_sd
feat_w2$summary$pitch_sd
feat_w3$summary$pitch_sd

```

### Spectral Domain

Spectral features of a soundwave reflect perturbances of a soundwave. Measures of spectral qualities of a soundave generally assess the amount of perturbance or periodicity of sound. Two such measure of perturbances can be directly extracted using the Harmonics-to-Noise ratio (HNR) and level of entropy of the soundwave from the `analyze( )` function of the `soundgen` package. Comparing the level of entropy and periodicity between the two commands confirms the moderately larger level of entropy and greater perturbances in the second compared to the first command.

```{r Mean entropy}
feat_cmd1$summary$entropy_mean
feat_cmd2$summary$entropy_mean
```

```{r Mean HNR}
feat_cmd1$summary$HNR_mean
feat_cmd2$summary$HNR_mean
```


# Summary and Conclusion

In this tutorial, we've explored the analysis of voice interactions with Amazon Alexa, focusing on a user's increasing frustration over multiple attempts. 

We've learned how to:

- Read and manipulate audio files using the tuneR and seewave packages
- Visualize sound data through oscillograms, spectrograms, and frequency plots
- Extract acoustic features across time, amplitude, frequency, and spectral domains
- Interpret vocal cues that indicate emotional states like frustration and anger

Our analysis revealed clear patterns of increased vocal intensity, higher pitch, longer voice breaks, and greater perturbances as the user became increasingly frustrated with the voice assistant. These techniques provide valuable insights for both research and practical applications in voice interface design and voice analytics.

## Key Takeaways

Our analysis of the Alexa interactions demonstrates several important insights about voice analytics:

**1. Voice features can effectively capture emotional states even without understanding the semantic content of speech**

**2. User frustration manifests through measurable acoustic patterns**:

- Higher fundamental frequency (pitch)
- Greater amplitude (loudness)
- Increased perturbances and entropy
- Longer duration and more pronounced voice breaks


## Looking Ahead

In the next chapter, we will expand our voice analytics pipeline in two important directions:

1. **Speech-to-Text Using LLMs**: We'll explore how to transcribe speech using Large Language Models, enabling us to analyze both the acoustic features and the semantic content of voice interactions.

2. **Automated Analysis with `voiceR`**: We'll introduce the `voiceR` package (available [here](https://cran.r-project.org/web/packages/voiceR/)), which streamlines the feature extraction and reporting process. We developed this package to provider researchers with an efficient workflow, allowing you to analyze voice data in batches more effectively.

I hope this first introduction to the field of voice analytics helped to expand your methodological skillset to analyze both the "how" (acoustic features) and the "what" (semantic content) of people's voice interactions.
