---
title: "Case Study: Does Anger Sound Different from Calmness?"
author: "Your Group Name"
format: 
  html:
    theme: flatly
    code-background: true
    code-copy: true
    toc: true
    toc-depth: 3
    toc-location: left
---


## Abstract 

This case study walks you through the **complete voice analytics pipeline** from our course on a new dataset. We **read** the audio, **visualize** it, **preprocess and visualize again**, **extract features** (acoustic features, embeddings, and some combination), and finally **use those features to predict** the speaker's emotional state. The code deliberately reuses the same functions and patterns as in our `GSERM_voiceanalyticsWorkshop_Day1-5.qmd` (`readWave()`, `oscillo()`, `spectro()`, `timer()`, `voiceR::autoExtract()`, `talk::talkEmbed()`), so you can map each step back to what we did in class.

We use the **RAVDESS** corpus (Livingstone & Russo, 2018), focusing on a simple arousal contrast — **Calm vs. Angry**. To keep the sentence content constant we use only the *"Kids are talking by the door"* recordings, giving a balanced **192 clips**: 24 actors × 2 emotions × 2 intensities × 2 repetitions, split evenly across sex (48 per sex × emotion cell). Every clip in the `9_ravdess_calm_angry/` folder is part of the analysis.

::: {.callout-note}
## Learning objectives
By the end of this case study, you should be able to:

1. Assemble a reproducible pipeline from raw audio to a predictive model on a multi-speaker corpus.
2. Read, visualize, and preprocess a batch of recordings and confirm each step visually.
3. Extract three complementary feature representations: `voiceR` acoustic features, `talk` embeddings, and their combination.
4. Fit and evaluate models with a **speaker-independent** train/test split, and read which features drive the prediction.
5. Discuss potential construct-validity and fairness issues before trusting the result.
:::

# Overview

Our question of this case is whether we separate calm from angry speech. The arousal account predicts anger raises pitch, loudness, and pitch variability and degrades voice quality (lower harmonics-to-noise ratio). As before, we simply follow the same steps as prevously: **read → visualize → preprocess → extract → predict**.

```{r setup, message=FALSE, warning=FALSE}
# Acoustics & audio (same core stack as in our GSERM workshop)
library(tuneR)
library(seewave)
library(soundgen)
library(voiceR)

# Data wrangling, modelling & validation
library(tidyverse)
library(glmnet)
library(pROC)
library(broom)

emo_dir <- "voiceData/9_ravdess_calm_angry"
```

---

# Reading the Data

## Reading the recordings from the folder

All 192 clips live in `voiceData/9_ravdess_calm_angry/`. Exactly as in the workshop, we `list.files()` the folder and read every recording directly into a named `wav_list` with `readWave()`.

```{r read-folder, warning=FALSE}
emo_files <- list.files(emo_dir, pattern = "\\.wav$")
length(emo_files)   # 192

# Read every file into a named list (workshop pattern; name = filename without .wav)
wav_list <- list()
for (i in seq_along(emo_files)) {
  name_key <- gsub("\\.wav$", "", emo_files[i])
  wav_list[[name_key]] <- readWave(file.path(emo_dir, emo_files[i]))
}

# Inspect two Wave objects to confirm
wav_list[[1]];wav_list[[192]]
#listen(wav_list[[1]])
#listen(wav_list[[192]])

```

## Parsing the design from filenames

RAVDESS encodes the experimental design in a 7-part filename, e.g. `03-01-05-01-01-01-12.wav`. As in the workshop, reading the data means recovering that metadata. We list **every file in the folder** and parse it with base R (the same `gsub()` logic we used on the `ID_Emotion_Bar` corpus, adapted to the `-` separator).

| Position | Field | Meaning |
|---|---|---|
| 3 | **Emotion** | 02 = calm, 05 = angry |
| 4 | Intensity | 01 = normal, 02 = strong |
| 5 | Statement | 01 = "Kids…" (the only one here) |
| 7 | **Actor** | 01–24 (odd = male, even = female) |

```{r parse-metadata}
audioDat <- data.frame(file = emo_files, stringsAsFactors = FALSE)
audioDat$emotion_code <- sapply(strsplit(audioDat$file, "-"), `[`, 3)
audioDat$actor        <- as.integer(gsub("\\.wav$", "", sapply(strsplit(audioDat$file, "-"), `[`, 7)))
audioDat$emotion      <- ifelse(audioDat$emotion_code == "02", "Calm", "Angry")
audioDat$sex          <- ifelse(audioDat$actor %% 2 == 0, "Female", "Male")

# The full folder is balanced across sex x emotion
table(audioDat$sex, audioDat$emotion)
```

---

::: {.callout-important}
## Your task from here on
The data is read in now in the previous steps. The rest of the pipeline is **yours to build** now. Work through the three sections below on your own, reusing the functions and patterns from our other workshop notebook. Each section lists ideas and directions — treat them more as ideas and initial suggestions but not as a checklist, and follow the threads you find most interesting!
:::

# Visualization & Pre-Processing

*Goal: get to know the raw signal, then clean it so later measurements describe speech rather than noise or silence.*

Ideas and directions to explore:

- Pick one **calm** and one **angry** clip from the *same* actor. Plot their **oscillograms** and **spectrograms** side by side (`oscillo()`, `spectro()` with `flim`, `osc = TRUE`, `dB = "max0"`). What differs — amplitude, where the energy sits in frequency?
- Because `oscillo()`/`spectro()` rescale each clip to its own maximum, side-by-side plots can hide a loudness difference. Try concatenating the two clips with `bind()` and plotting the combined wave on **one shared scale**. Does the shift become clearer?
- Detect the **speech onset** (and possibly the **offset**) automatically with `timer()`, and mark it on the oscillogram with `abline()`. Do the detected times match what you *hear*? Watch for low-level recording artifacts in the lead-in.
- Experiment with the `threshold` argument. What happens if it is too low (picks up noise) or too high (misses soft speech)? 
- **Trim** the leading/trailing silence (`extractWave()` from onset to offset, or `noSilence()`), then **re-plot** to confirm the edit did what you intended.
- Reflect: should you trim silence *before* extracting features? What could go wrong if you cut too aggressively?

```{r visualization-preprocessing}
# Your code here

```

# Feature Extraction

*Goal: turn each recording into numbers. Build (at least) two complementary representations and consider combining them.*

Ideas and directions to explore:

- **Acoustic features:** run `voiceR::autoExtract()` (or the `analyze` function from `soundgen`) on your `wav_list` with `extended = TRUE`. Think carefully about the `fileNamePattern` and `sep` — the RAVDESS names use `-` and numeric codes, so you will need to map the emotion code to `Condition` and the actor to `ID` (and mark irrelevant positions as `Null`). Recode `Condition` into `Calm`/`Angry`.
- Inspect the features by **feature family** (time, amplitude, frequency, spectral/voice-quality). Plot a few by emotion (e.g. `pitch_mean`, `loudness_mean`, `HNR_mean`, `pitch_sd`). Which families look most different for calm vs. angry?
- Check **data quality**: any missing values? constant or all-`NA` columns? How will you handle them before modelling?
- **Embeddings:** use `talk::talkEmbed()` to get a dense (768-dim) representation of each clip. This is slow one file at a time — consider a parallel loop / cluster as in the lecture
- **Combine:** join the acoustic features and the embeddings into one table (key on the filename). You now have three representations to compare: acoustic-only, embeddings-only, and combined.
- Reflect: which representation do you *expect* to work best for this arousal contrast, and why?

```{r feature-extraction}
# Your code here

```

# Modeling

*Goal: use the features to predict calm vs. angry.*

Ideas and directions to explore:

- **Fit a classifier** to predict emotion from the acoustic features — e.g. logistic regression (`glm`) or a penalized/lasso version (`glmnet`), which also handles many correlated features. 
- **Compare representations:** acoustic vs. embeddings vs. combined, using the *same* split and metric. Does the neural representation help? Does combining add anything?
- **What predicts anger?** For an interpretable model, read off the selected/important features. Relate them to the arousal account (pitch, loudness, spectral energy, voice quality). Careful with correlated features when interpreting coefficient signs.
- **Robustness / fairness:** does performance hold across subgroups — **female vs. male**, and **normal vs. strong** intensity? Where is the model weakest, and why might that be?
- **Discuss generalization issues:** Can we trust these results if we would apply this model in other contexts? Why vs. why not?

```{r modeling}
# Your code here

```

---

# References

- Livingstone SR, Russo FA (2018). The Ryerson Audio-Visual Database of Emotional Speech and Song (RAVDESS). *PLoS ONE* 13(5): e0196391. <https://doi.org/10.1371/journal.pone.0196391>. Data: <https://doi.org/10.5281/zenodo.1188976>.

---

# Course Instructor & Contact:
**Prof. Dr. Christian Hildebrand** <br>
Full Professor of Marketing Analytics <br>
Executive Director, Institute of Behavioral Science & Technology (IBT-HSG) <br> 
University of St. Gallen <br><br>
Email: [christian.hildebrand@unisg.ch](mailto:christian.hildebrand@unisg.ch) <br>
Web: [ibt.unisg.ch](https://ibt.unisg.ch) | Social: [LinkedIn Profile](https://www.linkedin.com/in/christian-hildebrand/)
